Performance comparison of Dask and Apache Spark on HPC systems \HL{for Neuroimaging}

### 2022-09-12 latex/paper.tex
[FIX] address comments + [FIX] grammar / syntax errors.

\section{Introduction}
The rise in data sharing coupled with improved data collection technologies leads neuroimaging into the Big Data era\cite{ALFAROALMAGRO2018400, van2014human, ConpPortal}.
While standard neuroimaging workflow engines, such as Nipype\cite{Nipype:11}, are well rounded to process standard compute-intensive pipelines,
they lack Big Data strategies (i.e., in-memory computing, data locality, and lazy-evaluation) to improve the performance of increasingly prevalent data-intensive pipelines.
A previous study\cite{8752675} noted the importance of in-memory computing and data locality to improve the performance of data-intensive pipelines.
This work studies how performance benchmarks \HL{of neuroimaging applications} generalize across Big Data engines in an HPC environment.
\HL{
	With the rising need for Big Data engines in neuroimaging and none commonly used
	in the field, it is important to evaluate performance of different engines.
}
In particular, we study in-depth the performance difference between Dask\cite{Dask:15} and Apache Spark\cite{Spark:16} for their suitability to process neuroimaging pipelines.
\HL{
	Our main contribution is to assess whether Spark or Dask has substantial performance benefits in processing neuroimaging pipelines in HPC systems.
}

@@

\HL{
	While our infrastructure is not of the scale of TOP500 HPC systems,
	it is hard to reserved multiple nodes concurrently on those infrastructure.
	Moreover, by using our infrastructure, we can assure that no external processes
	are influencing the benchmark for our experiements.
	Our system uses similar hardware as Compute Canada's B\'eluga cluster
	--- a representative cluster available to researchers in Canada --- and 
	replicates as close as possible its configuration.
	This lets us compare Spark and Dask in an isolated environement with the condition 
	of a realistic  allocation of resources, for neuroimaging applications, in an HPC system.
}

\HL{
	We configure Dask to have eight worker processes per node, each with eight threads.
	This is the default recommended configuration for the Dask SLURM cluster configuration~{\cite{DaskSLURMDoc}}.
	Similarly, we configured Spark to have eight executors per node (hereafter called workers for consistency with Dask), each with eight threads.
	When trying different configuration for the workers, we did not observe performance difference
	for our benchmark.
	Therefore we chose the recommended configuration from Dask.
}
Each worker was allocated \SI{31.5}{\gibi\byte} of memory.
For Spark, the JVM heap space uses 10\% of workers' memory, and the location of the log is set on an NFS-mounted directory.
The worker workspace for both engines is located on Lustre due to the limited fast storage space on the worker nodes.
Other Dask configurations are left to default.
A new Dask or Spark cluster was started and torn down for each experiment.
Dask \textit{2021.7.0} and Spark \textit{3.1.2} were used.

@@

\HL{
	We chose these dataset as they represent common dataset size for neuroimaging 
	imaging studies and for their free access.
}

\subsection{Applications}
This section describes the applications used to benchmark Dask and Spark.
We use three simple synthetic applications to have a deep understanding of the 
underlying behavior of the applications with simple I/O patterns, namely
\textit{Increment}, \textit{Multi-Increment}, and \textit{Histogram}.
Also, two neuroimaging applications are used to study more realistic 
applications, namely \textit{Kmeans} and \textit{BIDS App Example}{\cite{gorgolewski2017bids}}.
\HL{
	We chose these neuroimaging data processing applications to study the 
	effects from various Big Data characteristics.
	Table~{\ref{table:applications}} depicts the Big Data characteristics used in
	each application.
}

@@

\HL{
	To minimize the performance difference from implementation decisions, the 
	code base of the applications is common for both engines.
	Therefore, the performance observed comes directly from the engine scheduling and
	their APIs.
}

@@

\subsubsection{Kmeans}
For our third application, we apply Kmeans clustering to the voxel
intensities of the BigBrain image. We set the number of clusters to 3. The
application starts by reading the image blocks, combining all voxels in a
1-D array, and choosing initial centroids using the min, max, and
intermediate values. It assigns each voxel to its closest centroid and
updates each centroid with the average of its assigned voxels. It repeats
the assignment and updates steps for a configurable number of iterations.
Finally, the voxels of the image blocks are classified and written back to
the file system. Updating the centroids involves substantial data
communication between the workers.
The computation to find the closest centroids for each voxels scales with the number of centroids.
It can be computed in parallel at the cost of higher memory or sequentially to reduce memory usage.
We decided to compute the nearest centroids in parallel since we use only three centroids.
However, due to resources limitation on our infrastructure, the Kmeans experiments
are performed with only half of the BigBrain dataset ({\SI{303}{\gibi\byte}}).
\HL{We chose this applications for its real-life use case, large data shuffling between workers,
and requirements to keep all data in-memory.}

\subsubsection{BIDS App Example}
Our fourth application is the BIDS App Example, a neuroimaging pipeline to
measure brain volume from MRIs. For this application, we use the CoRR
dataset. The application extracts the brain volume of each participant
then computes the average for each group of participants. Unlike the other
applications, the BIDS App Example is a command-line executed in a Docker image
(bids/example on DockerHub). We converted the Docker image to a Singularity
image for use in HPC environments, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}
\HL{We chose this application for its real-life use case and it is containerized unlike our applications.}

\subsection{Experiments}
Table~\ref{table:parameters} shows the parameters that we varied
throughout the experiments. We varied (1) the number of nodes to assess
the scalability of the engines' scheduler, (2) the number of files \HL{to segment}
BigBrain to measure the effect of different I/O patterns and parallelization degrees, and (3) the
number of iterations to evaluate the effect of the number of tasks.
It should be noted that increasing the number of iterations for a given sleep 
delay also increases the total compute time.
\HL{
	The choice of parameters for our experiements was restrained by the datasets
	and resources available from the cluster.
	The unrestrained parameters were chosen to have an expected linear scaling 
	to facilitate the interpretation of the results.
}

@@

\section{Results}
\subsection{Increment}
Figure \ref{fig:increment_worker} depicts the total execution time spent in each code segment of the \textit{Increment} application.
The bars are broken down into Overhead, Read, Write, and other application-specific functions.
The mean of each bar is calculated from ten repetitions, with the error bar representing the standard deviation.
As expected, the compute time, from \textit{Increment}, remains the same when we vary the number of nodes.
However, the \textit{Read} and \textit{Write} time increases with the number of nodes for both engines.
\HL{
	With a large amount of worker threads accessing the Lustre file system
	concurrently, the disk bandiwdth or network are saturated and it create 
	an I/O bottleneck to access the Lustre file system.
}

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is overall similar.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
\HL{
	On the one hand, when multiple threads access Lustre concurrently they risk 
	of saturation the shared disk bandwidth, thus resulting in longer I/O time.
	On the other hand, when most threads are either idle or performing computations,
	it reduces the contention on the disk bandwidth resulting in shorter I/O time.
}
The only significant difference in total time is when two nodes are used.
This results in an approximately 3\% total runtime difference.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

\subsection{Kmeans}
\HL{
	When varying the number of workers, we observed that Dask is multifold faster
	than Spark, resulting in a $1.79\times$ to $2.71\times$ speed-up ratio on the makespan.
	The problem we observed was two fold.
	Firstly, the process ID of the workers' threads is not reported correctly and 
	leading to falsely excessive overhead for Spark.
	Secondly, the Spark workers were restarting due to lack of memory, leading 
	to task recomputation and having to reload data into memory.
}

@@

In a follow-up experiment, we further optimized our Kmeans implementations for both Dask and Spark to minimize the size of data kept in memory.
Figure~{\ref{fig:dask_kmeans_gantt}} and Figure~{\ref{fig:spark_kmeans_gantt}} show the Gantt chart for the execution
of this new experiments for Dask and Spark.
The x-axis shows the time in second while the y-axis depicts the workers, with 
one worker thread ID (Dask) and process ID (Spark) per line.
We represent each task on the Gantt chart with its associated function's color.
We again observe that Spark reports three times more processes than Dask threads.
\HL{
	It is unclear what is the cause for reporting extra Spark processes.
	However, with our analysis we found that both Dask and Spark use the same 
	amount of maximum concurrent processes or threads.
	Additionally, the is no significant difference in the average amount of 
	concurrent processes or threads between Dask and Spark.
}

@@

\subsection{Memory management}
% TODO remore reference to figure.
\HL{Our initial Kmeans results showe a significant difference between Dask and Spark.}
The difference in performance \HL{came} from Spark worker restarting due to running out of memory; thus, Spark \HL{needed} to recompute several tasks.
This is unexpected, given that Dask and Spark \HL{were} configured similarly and \HL{were} allocated the same amount of resources.
We think those issues arise from a combination of
(1) the reserved memory for the JVM heap space,
(2) the serialization of Python to Java kept in memory,
(3) Spark using processes compared to Dask using threads, thus limiting shared memory consumption,
and (4) different policies to persist data to disk when running out of memory.

@@

\HL{
	In this study, we observed that Dask has a larger startup overhead that Spark.
	These results are consistent with a previous study~{\cite{Mehta:17}}.
	However, other studies found the opposite~{\cite{8588652}} or that the difference was negligible~{\cite{8943502}}
}
Those apparent contractions most likely come from differences in infrastructure and configuration of the engines.

@@

\subsection{Recommendations}
\HL{
	The results from our experiments show that the performance difference between
	Dask and Spark is minimal, in general. 
	The main differences we observed are qualitative.
	We found that Dask integration with the scientific Python ecosystem made it
	easier to implement neuroimaging application.
	Spark configuration required to set multiple environment variables.
	In comparison, Dask is ready-to-go after an installation through PIP.
	Dask dashboard provides more information than Spark ones, making
	the debugging of Dask application easier and faster.
	Therefore, for the neuroimaging field, we recommend the use of Dask over Spark.
	However, in other fields with non-Python-based ecosystem, both Dask and spark
	would be equal candidate as a Big Data engine.
}

\HL{
	% TODO add caveats to avoid when using the engines.

}

\subsection{Limitations}
To make a fair comparison between Spark and Dask, we set both engine configurations
to be equivalent to the best of our capabilities, which comes with some limitations.
Firstly, both engines use a shared implementation for the algorithms.
This is done to keep the theoretical runtime equivalent; however, the chosen
implementation might not be optimal for either of the engines.
Secondly, we decided to use Lustre as a file system for both Spark and Dask.
Spark could benefit from using HDFS as a file system, but HDFS is uncommon in HPC.
\HL{Finally, our Lustre filesystem suffer from an I/O bottleneck coming from 
either the network or disks bandwidth.}
As explained in section~{\ref{subsec:overhead-tradeoff}}, the data transfer 
time is tightly correlated to the overhead of the application.
However, while reducing this bottleneck should improve the performance of 
both engines similarly, it can be hard to predict the actual behavior.

\HL{
	% TODO discuss UCX and other high performance networks.???

}

\HL{
	Dask and Spark both support GPUs, however we did not use applications that
	leverage them in our work.
	Although several neuroimaging applications use GPUs, only a few preprocessing pipelines use them.
	With preprocessing comprising the majority of compute time for neuroimaging
	workflow, we ommitted GPUs from this study.
	Moreover, the few preprocessing applications using GPUs are not commonly used
	due to their recent development.
}

>>>>

\section{Introduction}
The rise in data sharing coupled with improved data collection technologies leads neuroimaging into the Big Data era\cite{ALFAROALMAGRO2018400, van2014human, ConpPortal}.
While standard neuroimaging workflow engines, such as Nipype\cite{Nipype:11}, are well rounded to process standard compute-intensive pipelines,
they lack Big Data strategies (i.e., in-memory computing, data locality, and lazy-evaluation) to improve the performance of increasingly prevalent data-intensive pipelines.
A previous study\cite{8752675} noted the importance of in-memory computing and data locality to improve the performance of data-intensive pipelines.
This work studies how performance benchmarks \HL{of neuroimaging applications} generalize across Big Data engines in an HPC environment.
\HL{
	With the rising need for Big Data engines in neuroimaging and none commonly
	used in the field, evaluating the performance of different frameworks is essential.
}
In particular, we study in-depth the performance difference between Dask\cite{Dask:15} and Apache Spark\cite{Spark:16} for their suitability to process neuroimaging pipelines.
\HL{
	Our main contribution is to assess whether Spark or Dask has substantial performance benefits in processing neuroimaging pipelines in HPC systems.
}

@@

\HL{
	While our infrastructure is not of the scale of TOP500 HPC systems, it is 
	hard to reserve multiple nodes concurrently on those infrastructures.
	Moreover, by using our infrastructure, we can ensure that no external
	processes influence the benchmark for our experiments.
	Our system uses similar hardware as Compute Canada's B\'eluga cluster
	--- a representative cluster available to researchers in Canada --- and
	replicates as close as possible its configuration.
	This lets us compare Spark and Dask in an isolated environment with the
	condition of a realistic allocation of resources, for neuroimaging
	applications, in an HPC system.
}

\HL{
	We configure Dask to have eight worker processes per node, each with eight threads.
	This is the default recommended configuration for the Dask SLURM cluster 
	configuration~{\cite{DaskSLURMDoc}}.
	Similarly, we configured Spark to have eight executors per node
	(hereafter called workers for consistency with Dask), each with eight threads.
	However, when trying different configurations for the workers, we did not
	observe performance differences for our benchmark.
	Therefore we chose the recommended configuration from Dask.
}
Each worker was allocated \SI{31.5}{\gibi\byte} of memory.
For Spark, the JVM heap space uses 10\% of workers' memory, and the location of the log is set on an NFS-mounted directory.
The worker workspace for both engines is located on Lustre due to the limited fast storage space on the worker nodes.
Other Dask configurations are left to default.
A new Dask or Spark cluster was started and torn down for each experiment.
Dask \textit{2021.7.0} and Spark \textit{3.1.2} were used.

@@

\HL{
	We chose these datasets as they represent typical dataset sizes for neuroimaging studies and their free access.
}

\subsection{Applications}
This section describes the applications used to benchmark Dask and Spark.
We use three simple synthetic applications to have a deep understanding of the 
underlying behavior of the applications with simple I/O patterns, namely
\textit{Increment}, \textit{Multi-Increment}, and \textit{Histogram}.
Also, two neuroimaging applications are used to study more realistic 
applications, namely \textit{Kmeans} and \textit{BIDS App Example}{\cite{gorgolewski2017bids}}.
\HL{
	We chose these neuroimaging data processing applications to study the 
	effects of various Big Data characteristics.
	Table~{\ref{table:applications}} depicts the Big Data characteristics used in
	each application.
}

@@

\HL{
	To minimize the performance difference from implementation decisions, the
	code base of the applications is common for both engines.
	Therefore, the performance observed comes directly from the 
	engine scheduling and APIs.
}

@@

\subsubsection{Kmeans}
For our third application, we apply Kmeans clustering to the voxel
intensities of the BigBrain image. We set the number of clusters to 3. The
application starts by reading the image blocks, combining all voxels in a
1-D array, and choosing initial centroids using the min, max, and
intermediate values. It assigns each voxel to its closest centroid and
updates each centroid with the average of its assigned voxels. It repeats
the assignment and updates steps for a configurable number of iterations.
Finally, the voxels of the image blocks are classified and written back to
the file system. Updating the centroids involves substantial data
communication between the workers.
The computation to find the closest centroids for each voxels scales with the number of centroids.
It can be computed in parallel at the cost of higher memory or sequentially to reduce memory usage.
We decided to compute the nearest centroids in parallel since we use only three centroids.
However, due to resources limitation on our infrastructure, the Kmeans experiments
are performed with only half of the BigBrain dataset ({\SI{303}{\gibi\byte}}).
\HL{We chose this application for its real-life use case, large data shuffling between workers,
and requirements to keep all data in-memory.}

\subsubsection{BIDS App Example}
Our fourth application is the BIDS App Example, a neuroimaging pipeline to
measure brain volume from MRIs. For this application, we use the CoRR
dataset. The application extracts the brain volume of each participant
then computes the average for each group of participants. Unlike the other
applications, the BIDS App Example is a command-line executed in a Docker image
(bids/example on DockerHub). We converted the Docker image to a Singularity
image for use in HPC environments, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}
\HL{We chose this application for its real-life use case and it is containerized unlike our other applications.}

\subsection{Experiments}
Table~\ref{table:parameters} shows the parameters that we varied
throughout the experiments. We varied (1) the number of nodes to assess
the scalability of the engines' scheduler, (2) the number of files \HL{to segment}
BigBrain to measure the effect of different I/O patterns and parallelization degrees, and (3) the
number of iterations to evaluate the effect of the number of tasks.
It should be noted that increasing the number of iterations for a given sleep 
delay also increases the total compute time.
\HL{
	The choice of parameters for our experiments was restrained by the datasets
	and resources available from the cluster.
	The unrestrained parameters were chosen to have an expected linear scaling
	to facilitate the interpretation of the results.
}

@@

\section{Results}
\subsection{Increment}
Figure \ref{fig:increment_worker} depicts the total execution time spent in each code segment of the \textit{Increment} application.
The bars are broken down into Overhead, Read, Write, and other application-specific functions.
The mean of each bar is calculated from ten repetitions, with the error bar representing the standard deviation.
As expected, the compute time, from \textit{Increment}, remains the same when we vary the number of nodes.
However, the \textit{Read} and \textit{Write} time increases with the number of nodes for both engines.
\HL{
	With a large number of worker threads accessing the Lustre file system
	concurrently, the disk bandwidth or network is saturated, and it creates 
	an I/O bottleneck to access the Lustre file system.
}

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is overall similar.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
\HL{
	On the one hand, when multiple threads access Lustre concurrently, they risk 
	saturation of the shared disk bandwidth, thus resulting in longer I/O time.
	On the other hand, when most threads are either idle or performing computations,
	it reduces the contention on the disk bandwidth resulting in shorter I/O time.
}
The only significant difference in total time is when two nodes are used.
This results in an approximately 3\% total runtime difference.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

\subsection{Kmeans}
\HL{
	When varying the number of workers, we observed that Dask is multifold faster
	than Spark, resulting in a $1.79\times$ to $2.71\times$ speed-up ratio on the makespan.
	The problem we observed was twofold.
	Firstly, the process ID of the workers' threads is not reported correctly and 
	leading to falsely excessive overhead for Spark.
	Secondly, the Spark workers were restarting due to a lack of memory, leading 
	to task recomputation and having to reload data into memory.
}

@@

In a follow-up experiment, we further optimized our Kmeans implementations for both Dask and Spark to minimize the size of data kept in memory.
Figure~{\ref{fig:dask_kmeans_gantt}} and Figure~{\ref{fig:spark_kmeans_gantt}} show the Gantt chart for the execution
of this new experiments for Dask and Spark.
The x-axis shows the time in second while the y-axis depicts the workers, with 
one worker thread ID (Dask) and process ID (Spark) per line.
We represent each task on the Gantt chart with its associated function's color.
We again observe that Spark reports three times more processes than Dask threads.
\HL{
	It is unclear what the cause is for reporting extra Spark processes.
	However, with our analysis, we found that both Dask and Spark use the same 
	amount of maximum concurrent processes or threads.
	Additionally, the is no significant difference in the average amount of 
	concurrent processes or threads between Dask and Spark.
}

@@

\subsection{Memory management}
\HL{Our initial Kmeans results showed a significant difference between Dask and Spark.}
The difference in performance \HL{came} from Spark worker restarting due to running out of memory; thus, Spark \HL{needed} to recompute several tasks.
This is unexpected, given that Dask and Spark \HL{were} configured similarly and \HL{were} allocated the same amount of resources.
We think those issues \HL{arose} from a combination of
(1) the reserved memory for the JVM heap space,
(2) the serialization of Python to Java kept in memory,
(3) Spark using processes compared to Dask using threads, thus limiting shared memory consumption,
and (4) different policies to persist data to disk when running out of memory.

@@

\HL{
	In this study, we observed that Dask has a larger startup overhead than Spark.
	These results are consistent with a previous study~{\cite{Mehta:17}}.
	However, other studies found the opposite~{\cite{8588652}} or that the difference was negligible~{\cite{8943502}}.
}
Those apparent contractions most likely come from differences in infrastructure and configuration of the engines.

@@

\HL{
	Another study~{\cite{Slaughter2020-ql}} developed a parametrized benchmark framework
	to evaluate the performance of different engines, including Spark and Dask.
	On the one hand, the framework reduces the implementation effort and helps with
	comparing frameworks fairly.
	On the other hand, it is hard to characterize neuroimaging realistically
	applications using their framework since the runtime of these applications can vary
	greatly depending on the subject process, and the framework lacks kernels with
	a variable amount of compute-intensity.
	Due to these limitations, we decided to implement the benchmarked applications
	directly into Spark and Dask.
}

\subsection{Recommendations}
\HL{
	The results from our experiments show that the performance difference between
	Dask and Spark are minimal, in general. 
	The main differences we observed are qualitative.
	First, we found that Dask integration with the scientific Python ecosystem made it
	easier to implement neuroimaging applications.
	Spark configuration requires setting multiple environment variables.
	In comparison, Dask is ready to go after an installation through PIP.
	In addition, Dask dashboard provides more information than Spark ones, making
	the debugging of Dask applications easier and faster.
	Therefore, for the neuroimaging field, we recommend using Dask over Spark.
	However, in other fields with non-Python-based ecosystems, both Dask and Spark
	would be equal candidates as Big Data engines.
}

\HL{
	While the engines showed similar performance, they both come with their caveats that
	can drastically affect their performance.
	For Dask, the user should be aware of the Python GIL limitations.
	For Python code, multiples worker processes should be used while for code releasing the
	GIL, a balance between processes and thread is recommended.
	For Spark, we experienced severe performance issues due to workers running out of memory.
	Users should carefully consider their Spark worker configurations to prevent similar issues.
}

\subsection{Limitations}
To make a fair comparison between Spark and Dask, we set both engine configurations
to be equivalent to the best of our capabilities, which comes with some limitations.
Firstly, both engines use a shared implementation for the algorithms.
This is done to keep the theoretical runtime equivalent; however, the chosen
implementation might not be optimal for either of the engines.
Secondly, we decided to use Lustre as a file system for both Spark and Dask.
Spark could benefit from using HDFS as a file system, but HDFS is uncommon in HPC.
\HL{Finally, our Lustre filesystem suffers from an I/O bottleneck coming from 
either the network or disk bandwidth.}
As explained in section~{\ref{subsec:overhead-tradeoff}}, the data transfer 
time is tightly correlated to the overhead of the application.
However, while reducing this bottleneck should improve the performance of 
both engines similarly, it can be hard to predict the actual behavior.

\HL{
	While both Dask and Spark support various high-performance network strategies
    such as UCX, we focused on the standpoint of neuroscientists that would 
    commonly use the frameworks mostly out of the box.
    With higher network throughput, the overall performance of the engines could differ.
    Further work would be required to compare Dask and Spark using different high-performance
    network protocols.
}

\HL{
	Dask and Spark both support GPUs; however, we did not use applications that
    leverage them in our work.
    Although several neuroimaging applications use GPUs, only a few preprocessing pipelines use them.
    Therefore, we omitted GPUs from this study because preprocessing comprises the majority of computing time for neuroimaging
    workflow.
    Moreover, the few preprocessing applications using GPUs are not commonly used
    due to their recent development.
}
